---
title: "Case Study 1: The Butterfly Ballot Effect: Investigating the Impact Butterfly Ballots had on Voting in the 2000 US Presidential Election in Florida Counties"
author: "Abby Hunt and Tajhini Brown"
date: "March 5, 2025"
output: 
  bookdown::pdf_document2: default
format: 
  pdf:
    keep-tex: true
    include-in-header: 
       text: |
         \usepackage{fvextra}
         \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
         \DefineVerbatimEnvironment{OutputCode}{Verbatim}{breaklines,commandchars=\\\{\}}
    geometry: 
      - left=1in
      - right=1in
      - top=1in
      - bottom=1in 
---

## Introduction

The 2000 US presidential election was one of the closest in history. The race was between the Republican nominee George W. Bush and Al Gore, the incumbent Democratic vice president. Florida was the tie-breaker state. Television networks’ predictions oscillated between both candidates as victors as votes were counted. At the end of counting, Bush was leading by only 1,738 votes. The narrow margin called for an immediate recount. When votes were recounted, it was revealed that Bush had actually led by less than 400 votes. This caused Democratic voters in Palm Beach County to complain about the design of the ballot as it had a “butterfly” layout, and that they had accidentally voted for the wrong candidate, Pat Buchanan.

Using statistical analysis, this case study seeks to suggest answers to the number of votes Buchanan could have reasonably expected to receive given a more user-friendly ballot. Using a linear regression model, with Buchanan votes as a function of Bush votes, this case study seeks to determine the predicted number of votes for Buchanan in Palm Beach County, assuming that the relationship between Bush votes and Buchanan votes was the same in Palm Beach as in the other Florida counties. This case study also aims to determine how many mis-cast votes there were since some voters erroneously voted Buchanan instead of Gore.

## Description of Data and Exploratory Data Analysis

The dataset originates from the Statistical Sleuth data library. It consists of all total votes for Bush and Buchanan in the US 2000 presidential election for each county in Florida, which is 67 in total. The unit of observation is a vote cast in the United States 2000 presidential election. As Palm Beach County was a major outlier, it was removed from the data set to improve model accuracy. This brought the total observations to 66.

```{r}
#| message: FALSE
#| warning: FALSE
#| echo: FALSE
#| label: election-summary-statistics
#| tbl-cap: "United States 2000 Presidential Election Florida County Votes Summary Statistics "

library(tidyverse)
library(Sleuth2)
library(broom)        
library(kableExtra)   
library(ggplot2)
library(knitr)

# Loading the data for case study one
election <- Sleuth2::ex0825

# Creating a second dataset with Palm Beach County excluded
election_wo_pb <- election |> filter(County != "Palm Beach")
election_wo_pb_summary_df <- election_wo_pb |>
  select(Bush2000, Buchanan2000) 

sapply(election_wo_pb_summary_df, summary, digits=5) |>
  kable()
```

Based on Table 1, the minimum number of votes cast for Bush in any county was 1316, and the maximum was 289,456. Buchanan had a minimum of 9 votes cast in his favour in any county during the election and 3407. Bush’s median number of votes was significantly higher than Buchanan's at 20,196, while Buchanan had a median of 114. Bush had a mean of 43,356 votes, and Buchanan had a mean of 258.5 votes, indicating the typical total votes for a county. As we can see from the summary statistics, Bush consistently had more votes than Buchanan in the 2000 Presidential election.

The mean number of votes for both candidates is larger than the median number of votes, indicating an inflation of the mean by outlier total county votes. There are fewer outliers affecting the mean of Buchanan votes than that of Bush votes.

```{r}
#| message: FALSE
#| warning: FALSE
#| echo: FALSE
#| label: eda-histogram
#| fig-cap: "Histogram Showing Distribution of Bush and Buchanan Votes"

tidy_election_wo_pb <- election_wo_pb |>
  select(Buchanan2000, Bush2000)|>
  stack()

ggplot(tidy_election_wo_pb, aes(x=values, fill = ind)) +
  geom_histogram(bins = 20)+
  theme(legend.position="top")+
  labs(x = "Votes", y= "Count", fill = "Candidates")+
  scale_fill_manual(values=c("#A020F0", "#FF0000")) + 
  facet_wrap(~ind)
```

Figure 1 shows a considerable difference in the distribution of Buchanan votes compared to Bush votes. While both are right-skewed, the degree of skew on the plot of Buchanan votes is noticeably larger. Each county's total votes for Buchanan is less than 100,000. Total county votes for Bush ranged between the low hundred thousands and 200,000. The distribution indicates outlier total votes of around 300,000. The variation in the number of Bush votes across the counties of Florida is greater than that across Buchanan votes. To improve the skew of the distribution, a natural logarithmic transformation can be applied to both variables.

```{r}
#| message: FALSE
#| warning: FALSE
#| echo: FALSE
#| label: eda-histogram-log
#| fig-cap: "Histogram Showing Distribution of Bush and Buchanan Votes Under a Natural Logarithmic Transformation"
ggplot(tidy_election_wo_pb, aes(x=log(values), fill = ind)) +
  geom_histogram(bins = 20)+
  theme(legend.position="top")+
  labs(x = "log(Votes)", y= "Count", fill = "Candidates")+
  scale_fill_manual(values=c("#A020F0", "#FF0000")) + 
  facet_wrap(~ind)
```

Figure 2 shows a clearer image of the distribution of Buchanan's votes by reducing the skew. The histogram of Buchanan's votes shows a unimodal distribution, indicating a singular most frequent number of votes. The distribution of Bush votes shows a bimodal distribution, indicating two groups of votes with the highest frequency.

```{r}
#| message: FALSE
#| warning: FALSE
#| echo: FALSE
#| label: eda-scatterplot
#| fig-cap: "Scatterplot Showing Buchanan Votes against Bush Votes"
election_wo_pb |> ggplot(aes(x = Bush2000, y = Buchanan2000)) + 
  geom_point() + 
  labs(title ="Association Between Votes for Bush and Votes for Buchanan.", x = "Bush Votes", y = "Buchanan Votes")
```

We observe from Figure 3 that most of the Buchanan total county votes lie between 0 and 500, while most of the total county votes for Bush lie between 0 and 100,000.

```{r}
#| message: FALSE
#| warning: FALSE
#| echo: FALSE
#| label: eda-scatterplot-log
#| fig-cap: "Scatterplot Showing Buchanan Votes against Bush Votes Under a Natural Logarithmic Transformation"
ggplot(election_wo_pb, aes(y = Buchanan2000, x = Bush2000)) + 
  geom_point() + 
  labs(title ="Association Between Natural Logarithmic Transformation Votes for Bush and Votes for Buchanan.", x = "log(Bush Votes)", y = "log(Buchanan Votes)")+
  scale_x_continuous(trans = "log") + 
  scale_y_continuous(trans = "log")
```

We observe from Figure 4 a similar distribution of Buchanan and Bush Votes on a logarithmic scale.

## Statistical Methods and Results

To estimate the relationship between the quantitative outcome variable, number of Buchanan votes and the quantitative explanatory variable, the number of Bush votes in the 2000 election, a simple linear regression model was employed. Using a linear regression model, we can also predict a more reasonable point estimate for Buchanan votes based on this relationship. Through intervals around the model's predicated point estimate, we can also determine a plausible number of votes that should have been in Gore. The model expresses Buchanan votes as a function of Bush votes. To ensure the model is as appropriate for the data as the plot in Figure 3 shows a violation of the equal variance condition for simple linear regression, transformations are applied to both the explanatory and response variables.

###### Population Model for the Mean Buchanan Votes as a Function of Bush Votes

$$ E[log(Buchanan2000)|log(Bush2000)] = \beta_0 + \beta_1(log(Bush2000)), $$

where *log(Buchanan2000)* is the natural logarithm of the number of votes Pat Buchanan received in the 2000 United States presidential election and *log(Bush2000)* is the natural logarithm of the number of votes George W. Bush received in the 2000 United States presidential election.

###### The Fitted Model for the Mean

$$ \widehat{log({y_i})} = \widehat\beta_0 + \widehat\beta_1log(x_i) $$

Estimates for the intercept coefficient, $\widehat\beta_0$ and slope, $\widehat\beta_1$ values were determined by fitting the model to the data.

```{r}
#| message: FALSE
#| warning: FALSE
#| echo: FALSE
#| label: estimate-tbl
#| tbl-cap: "Table showing estimates of the model Parameters and their accompanying standard errors"
# Fit a linear regression model to predict Buchanan2000 votes based on Bush2000 votes
election_wo_pb_lm <- lm(log(Buchanan2000) ~ log(Bush2000), data = election_wo_pb)

# Representing the regression table as a dataframe (i.e., tidying the summary() output)
election_wo_pb_lm_table <- summary(election_wo_pb_lm)$coefficients

# Creating a nicely formatted table from the dataframe using the kable package
#   You can find more information about this package here: https://haozhu233.github.io/kableExtra/awesome_table_in_pdf.pdf
election_wo_pb_lm_table |> kbl(col.names = c("Estimate", "Standard Error", "t value", "Pr(>|t|)"), align = "c", booktabs = T, linesep="", digits = c(2, 2, 2, 4)) |> kable_classic(full_width = F, latex_options = c("HOLD_position"))
```

Based on the model estimates summary table, $\widehat\beta_0$ is -2.34, and $\widehat\beta_1$ is 0.73. The slope indicates that there is a positive relationship between the number of total county votes Bush received and the number of total county votes Buchanan received.

###### The Fitted Model for the Mean Buchanan Votes as a Function of Bush Votes

$$ \widehat{log({y_i})} = -2.34 + 0.73(log(x_i)) $$

A doubling of the number of total votes Bush receives in a county is associated with a multiplicative change of $2^{0.73}$ in the median of total Buchanan votes in a county.

To predict a value for the number of total county votes Buchanan could have reasonably expected to receive in Palm Beach County given a more user-friendly ballot, we used the fitted regression line and the number of votes Bush received in that county to calculate it. We also obtained a reasonable range for this value by using a 95% prediction interval.

Based on our fitted regression line, Buchanan's total number of votes in Palm County is more reasonably expected to be 592 based on the number of votes Bush received in the same county: 152,846 votes.

We are 95% confident that the number of votes for Buchanan for Palm County total votes with Bush total county votes at 152,846 is between 251 and 1399.

## Discussion

## R Appendix

```{r}
#| message: FALSE
#| warning: FALSE

# Loading necessary packages
library(tidyverse)
library(Sleuth2)
library(broom)        
library(kableExtra)   
library(ggplot2)
library(knitr)

# Loading the data for case study one
election <- Sleuth2::ex0825

# Creating a second dataset with Palm Beach County excluded
election_wo_pb <- election |> filter(County != "Palm Beach")

# Creating a scatterplot to visualize the relationship between Bush and Buchanan Votes
election_wo_pb |> ggplot(aes(x = Bush2000, y = Buchanan2000)) + geom_point() + ggtitle("Association between votes for Bush and votes for Buchanan.")

# Fit a linear regression model to predict Buchanan2000 votes based on Bush2000 votes
election_lm <- lm(Buchanan2000 ~ Bush2000, data = election_wo_pb)

# Representing the regression table as a dataframe (i.e., tidying the summary() output)
election_lm_table <- summary(election_lm)$coefficients

# Creating a nicely formatted table from the dataframe using the kable package
#   You can find more information about this package here: https://haozhu233.github.io/kableExtra/awesome_table_in_pdf.pdf
election_lm_table |> kbl(col.names = c("Estimate", "Standard Error", "t value", "Pr(>|t|)"), align = "c", booktabs = T, linesep="", digits = c(2, 2, 2, 4)) |> kable_classic(full_width = F, latex_options = c("HOLD_position"))

# Creating a vector of the number of Palm Beach votes for Bush 
election_pb_bush <- election |> 
              filter(County == "Palm Beach")|>
              pull(Bush2000)
election_pb_bush

#Use the fitted regression model to predict Buchanan2000 votes for Palm Beach County

new_election <- data.frame(Bush2000 = election_pb_bush)
palm_beach_predictions <- election_lm |> augment(newdata = new_election, 
                          interval = "prediction", 
                          conf.level = 0.95)

# Viewing the prediction interval
palm_beach_predictions  

# Fit a linear regression model using a double log transformation
model_double_log <- lm(log(Buchanan2000) ~ log(Bush2000), data = election_wo_pb)

# View the summary of the model to see the regression results
summary(model_double_log)

# Representing the regression table as a dataframe (i.e., tidying the summary() output)
model_double_log_table <- summary(model_double_log)$coefficients

# Creating a nicely formatted table from the dataframe using the kable package
#   You can find more information about this package here: https://haozhu233.github.io/kableExtra/awesome_table_in_pdf.pdf
model_double_log_table |> kbl(col.names = c("Estimate", "Standard Error", "t value", "Pr(>|t|)"), align = "c", booktabs = T, linesep="", digits = c(2, 2, 2, 4)) |> kable_classic(full_width = F, latex_options = c("HOLD_position"))

# Create a scatter plot of the log-transformed votes for Bush and Buchanan with a regression line
ggplot(election_wo_pb, aes(y = Buchanan2000, x = Bush2000)) + 
  geom_point() +
  geom_smooth(method=lm, se=FALSE, formula = y~x) +
  scale_x_continuous(trans = "log") + 
  scale_y_continuous(trans = "log")

# Generate predictions using the double-log regression model for new data (palm_beach_data),
new_election <- data.frame(Bush2000 = election_pb_bush)
palm_beach_predictions <- model_double_log |> augment(newdata = new_election, 
                          interval = "prediction", 
                          conf.level = 0.95)
palm_beach_predictions
# Providing both the predicted values and a 95% prediction interval

# Exponentiate the predictions (since we applied log transformation during modeling),
# to return the predicted Buchanan votes in the original scale (non-log scale).
pred_buchanan_double_log <- exp(palm_beach_predictions$.fitted)
# Display the transformed predictions (original scale of Buchanan votes)
pred_buchanan_double_log 

# Basic summary statistics for Bush and Buchanan votes
summary(election$Bush)
summary(election$Buchanan)

election_summary_df <- election |>
  select(Bush2000, Buchanan2000) 

sapply(election_summary_df, summary, digits=5) |>
  kable()
  

```
