---
title: "Case Study 1: The Butterfly Ballot Effect: Investigating the Impact Butterfly Ballots had on Voting in the 2000 US Presidential Election in Florida Counties"
author: "Abby Hunt and Tajhini Brown"
date: "March 5, 2025"
output: 
  bookdown::pdf_document2: default
format: 
  pdf:
    keep-tex: true
    include-in-header: 
       text: |
         \usepackage{fvextra}
         \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
         \DefineVerbatimEnvironment{OutputCode}{Verbatim}{breaklines,commandchars=\\\{\}}
    geometry: 
      - left=1in
      - right=1in
      - top=1in
      - bottom=1in 
---

## Introduction

The 2000 US presidential election was one of the closest in history. The race was between the Republican nominee George W. Bush and Al Gore, the incumbent Democratic vice president. Florida was the tie-breaker state. Television networks’ predictions oscillated between both candidates as victors as votes were counted. At the end of counting, Bush was leading by only 1,738 votes. The narrow margin called for an immediate recount. When votes were recounted, it was revealed that Bush had actually led by less than 400 votes. This caused Democratic voters in Palm Beach County to complain about the design of the ballot as it had a “butterfly” layout, and that they had accidentally voted for the wrong candidate, Pat Buchanan.

Using statistical analysis, this case study seeks to suggest answers to the number of votes Buchanan could have reasonably expected to receive given a more user-friendly ballot. Using a linear regression model, with Buchanan votes as a function of Bush votes, this case study seeks to determine the predicted number of votes for Buchanan in Palm Beach County, assuming that the relationship between Bush votes and Buchanan votes was the same in Palm Beach as in the other Florida counties. This case study also aims to determine how many mis-cast votes there were since some voters erroneously voted Buchanan instead of Gore.

## Description of Data and Exploratory Data Analysis

The dataset originates from the Statistical Sleuth data library. It consists of all total votes for Bush and Buchanan in the US 2000 presidential election for each county in Florida, which is 67 in total. The unit of observation is the total number of votes for a county in Florida in the United States 2000 presidential election. As Palm Beach County was a major outlier, it was removed from the data set to improve model accuracy. This brought the total observations to 66.

```{r}
#| message: FALSE
#| warning: FALSE
#| echo: FALSE
#| label: election-summary-statistics
#| tbl-cap: "United States 2000 Presidential Election Florida County Votes Summary Statistics "

library(tidyverse)
library(Sleuth2)
library(broom)        
library(kableExtra)   
library(ggplot2)
library(knitr)
libary(dplyr)

# Loading the data for case study one
election <- Sleuth2::ex0825

# Creating a second dataset with Palm Beach County excluded
election_wo_pb <- election |> filter(County != "Palm Beach")
election_wo_pb_summary_df <- election_wo_pb |>
  select(Bush2000, Buchanan2000) 

sapply(election_wo_pb_summary_df, summary, digits=5) |>
  kable()
```

Based on Table 1, the minimum number of votes cast for Bush in any county was 1316, and the maximum was 289,456. Buchanan had a minimum of 9 votes cast in his favor in any county during the election and 3407. Bush’s median number of votes was significantly higher than Buchanan's at 20,196, while Buchanan had a median of 114. Bush had a mean of 43,356 votes, and Buchanan had a mean of 258.5 votes, indicating the typical total votes for a county. As we can see from the summary statistics, Bush consistently had more votes than Buchanan in the 2000 Presidential election.

Based on Table 1, the minimum number of votes cast for Bush in any county was 1316, and the maximum was 289,456. Buchanan had a minimum of 9 votes cast in his favour in any county during the election and 3407. Bush’s median number of votes was significantly higher than Buchanan's at 20,196, while Buchanan had a median of 114. Bush had a mean of 43,356 votes, and Buchanan had a mean of 258.5 votes, indicating the typical total votes for a county. As we can see from the summary statistics, Bush consistently had more votes than Buchanan in the 2000 Presidential election.

The mean number of votes for both candidates is larger than the median number of votes, indicating an inflation of the mean by outlier total county votes. There are fewer outliers affecting the mean of Buchanan votes than that of Bush votes.

```{r}
#| message: FALSE
#| warning: FALSE
#| echo: FALSE
#| label: eda-histogram
#| fig-cap: "Histogram Showing Distribution of Bush and Buchanan Votes"

tidy_election_wo_pb <- election_wo_pb |>
  select(Buchanan2000, Bush2000)|>
  stack()

ggplot(tidy_election_wo_pb, aes(x=values, fill = ind)) +
  geom_histogram(bins = 20)+
  theme(legend.position="top")+
  labs(x = "Votes", y= "Count", fill = "Candidates")+
  scale_fill_manual(values=c("#A020F0", "#FF0000")) + 
  facet_wrap(~ind)
```

Figure 1 shows a considerable difference in the distribution of Buchanan votes compared to Bush votes. While both are right-skewed, the degree of skew on the plot of Buchanan votes is noticeably larger. Each county's total votes for Buchanan is less than 100,000. Total county votes for Bush ranged between the low hundred thousands and 200,000. The distribution indicates outlier total votes of around 300,000. The variation in the number of Bush votes across the counties of Florida is greater than that across Buchanan votes. To improve the skew of the distribution, a natural logarithmic transformation can be applied to both variables.

```{r}
#| message: FALSE
#| warning: FALSE
#| echo: FALSE
#| label: eda-histogram-log
#| fig-cap: "Histogram Showing Distribution of Bush and Buchanan Votes Under a Natural Logarithmic Transformation"
ggplot(tidy_election_wo_pb, aes(x=log(values), fill = ind)) +
  geom_histogram(bins = 20)+
  theme(legend.position="top")+
  labs(x = "log(Votes)", y= "Count", fill = "Candidates")+
  scale_fill_manual(values=c("#A020F0", "#FF0000")) + 
  facet_wrap(~ind)
```

Figure 2 shows a clearer image of the distribution of Buchanan's votes by reducing the skew. The histogram of Buchanan's votes shows a unimodal distribution, indicating a singular most frequent number of votes. The distribution of Bush votes shows a bimodal distribution, indicating two groups of votes with the highest frequency.

```{r}
#| message: FALSE
#| warning: FALSE
#| echo: FALSE
#| label: eda-scatterplot
#| fig-cap: "Scatterplot Showing Buchanan Votes against Bush Votes"
election_wo_pb |> ggplot(aes(x = Bush2000, y = Buchanan2000)) + 
  geom_point() + 
  labs(title ="Association Between Votes for Bush and Votes for Buchanan.", x = "Bush Votes", y = "Buchanan Votes")
```

We observe from Figure 3 that most of the Buchanan total county votes lie between 0 and 500, while most of the total county votes for Bush lie between 0 and 100,000.

```{r}
#| message: FALSE
#| warning: FALSE
#| echo: FALSE
#| label: eda-scatterplot-log
#| fig-cap: "Scatterplot Showing Buchanan Votes against Bush Votes Under a Natural Logarithmic Transformation"
ggplot(election_wo_pb, aes(y = Buchanan2000, x = Bush2000)) + 
  geom_point() + 
  labs(title ="Association Between Natural Logarithmic Transformation Votes for Bush and Votes for Buchanan.", x = "log(Bush Votes)", y = "log(Buchanan Votes)")+
  scale_x_continuous(trans = "log") + 
  scale_y_continuous(trans = "log")
```

We observe from Figure 4 a similar distribution of Buchanan and Bush Votes on a natural logarithmic scale.

## Statistical Methods and Results

To estimate the relationship between the quantitative outcome variable, number of Buchanan votes and the quantitative explanatory variable, the number of Bush votes in the 2000 election, a simple linear regression model was employed. Using a linear regression model, we can also predict a more reasonable point estimate for Buchanan votes based on this relationship. Through intervals around the model's predicated point estimate, we can also determine a plausible number of votes that should have been in Gore. The model expresses Buchanan votes as a function of Bush votes. To ensure the model is as appropriate for the data as the plot in Figure 3 shows a violation of the equal variance condition for simple linear regression, transformations are applied to both the explanatory and response variables.

###### Population Model for the Mean Buchanan Votes as a Function of Bush Votes

$$ E[log(Buchanan2000)|log(Bush2000)] = \beta_0 + \beta_1(log(Bush2000)), $$

where *log(Buchanan2000)* is the natural logarithm of the number of votes Pat Buchanan received in the 2000 United States presidential election and *log(Bush2000)* is the natural logarithm of the number of votes George W. Bush received in the 2000 United States presidential election.

###### The Fitted Model for the Mean

$$ \widehat{log({y_i})} = \widehat\beta_0 + \widehat\beta_1log(x_i) $$

Estimates for the intercept coefficient, $\widehat\beta_0$ and slope, $\widehat\beta_1$ , were determined by fitting the model to the data.

```{r}
#| message: FALSE
#| warning: FALSE
#| echo: FALSE
#| label: estimate-tbl
#| tbl-cap: "Table showing estimates of the model Parameters and their accompanying standard errors"
# Fit a linear regression model to predict Buchanan2000 votes based on Bush2000 votes
election_wo_pb_lm <- lm(log(Buchanan2000) ~ log(Bush2000), data = election_wo_pb)

# Representing the regression table as a dataframe (i.e., tidying the summary() output)
election_wo_pb_lm_table <- summary(election_wo_pb_lm)$coefficients

# Creating a nicely formatted table from the dataframe using the kable package
#   You can find more information about this package here: https://haozhu233.github.io/kableExtra/awesome_table_in_pdf.pdf
election_wo_pb_lm_table |> kbl(col.names = c("Estimate", "Standard Error", "t value", "Pr(>|t|)"), align = "c", booktabs = T, linesep="", digits = c(2, 2, 2, 4)) |> kable_classic(full_width = F, latex_options = c("HOLD_position"))
```

Based on the model estimates summary table, $\widehat\beta_0$ is -2.34, and $\widehat\beta_1$ is 0.73. The slope indicates that there is a positive relationship between the number of total county votes Bush received and the number of total county votes Buchanan received.

###### The Fitted Model for the Mean Buchanan Votes as a Function of Bush Votes

$$ \widehat{log({y_i})} = -2.34 + 0.73(log(x_i)) $$

A doubling of the number of total votes Bush receives in a county is associated with a multiplicative change of $2^{0.73}$ in the median of total Buchanan votes in a county.

To predict a value for the number of total county votes Buchanan could have reasonably expected to receive in Palm Beach County given a more user-friendly ballot, we used the fitted regression line and the number of votes Bush received in that county to calculate it. We also obtained a reasonable range for this value by using a 95% prediction interval.

Based on our fitted regression line, Buchanan's total number of votes in Palm County is more reasonably expected to be 592 based on the number of votes Bush received in the same county: 152,846 votes.

We are 95% confident that the number of votes for Buchanan for Palm County total votes with Bush total county votes at 152,846 is between 251 and 1399.

Given that the number of votes Buchanan received in Palm Beach County was 3407, we can conclude that there were between 3156 and 2008 mis-cast votes that should have gone to Gore.

## Discussion

Democracy is the foundation of the United States. Every four years, voters come out to exercise their constitutional right to determine the head of the government that represents them. The importance of the presidential election must be underscored.

In the first presedential election of the new millennium, a new ballot in Florida and close election shook this foundation. This case study sought to answer questions about an alternate voting outcome provided that the problems that arose did not exist.

This case study sought to suggest answers to the number of votes Buchanan could have reasonably expected to receive given a more user-friendly ballot. It sought to determine the predicted number of votes for Buchanan in Palm Beach County, assuming that the relationship between Bush votes and Buchanan votes was the same in Palm Beach as in the other Florida counties. This case study also sought to determine how many mis-cast votes there were since some voters erroneously voted Buchanan instead of Gore.

Through our statistical analysis using a simple linear regression model fit to the data, we determined that given that the ballot that was used in all other Florida counties was used in Palm Beach, Buchanan could have expected to receive 592 votes instead of the 3407 he received. After constructing a prediction interval for these values, his predicted total for Palm Beach County fell between 251 and 1399. Taking this prediction range as a reasonable prediction, there may have been between 3156 and 2008 mis-cast votes that should have gone to Gore.

While Bush would have still won this county in terms of total votes, it is important to identify flawed counts in important elections that have the power to affect people's lived experiences and the reasons for them to put in place preventative measures.

Since we removed Palm Beach, our analysis is based on the typical voting patterns in the other counties. This means that our model is more generalized to counties with more standard ballot designs and voting conditions. While Palm Beach's issues were likely the most extreme, other counties might have had smaller, unquantified ballot design flaws, voter confusion, or other irregularities that could still influence voting behavior. Our model does not capture any similar but less obvious problems in other counties.

## R Appendix

**Assessing the Linear Regression Model**

```{r}
#| message: FALSE
#| warning: FALSE
#| echo: FALSE
election_wo_pb_lm <- lm(Buchanan2000 ~ Bush2000, data = election_wo_pb)
plot(election_wo_pb_lm)
```

From the plots, the model violates the linearity and equal variance conditions. It also violates the normality condition, as the Q-Q plot shows many extreme residuals. Since the data is clustered by geography (county in Florida), the model violates the independence condition.

**Assessing the Linear Regression Model Under Natural Logarithmic Transformation of Response and Explanatory Variables**

```{r}
#| message: FALSE
#| warning: FALSE
#| echo: FALSE
election_wo_pb_lm_log <- lm(log(Buchanan2000) ~ log(Bush2000), data = election_wo_pb)
plot(election_wo_pb_lm_log)
```

From the plots, from the residuals vs fitted plot, we can observe that the natural logarithmic transformation improved the model, and it now meets the linearity and homoscedasticity conditions. While the Q-Q Residuals plot did increase the normality, there are still many extreme residuals to the left of the plot. This model also does not meet the independence condition.

```{r}
#| message: FALSE
#| warning: FALSE
#| output: FALSE

library(tidyverse)
library(Sleuth2)
library(broom)        
library(kableExtra)   
library(ggplot2)
library(knitr)
library(dplyr)

# Loading the data for case study one
election <- Sleuth2::ex0825

# Creating a second dataset with Palm Beach County excluded
election_wo_pb <- election |> filter(County != "Palm Beach")
election_wo_pb_summary_df <- election_wo_pb |>
  select(Bush2000, Buchanan2000) 

# Creating and formatting a table of summary statistics
sapply(election_wo_pb_summary_df, summary, digits=5) |>
  kable()

# Data wrangling the election dataset to create histograms
tidy_election_wo_pb <- election_wo_pb |>
  select(Buchanan2000, Bush2000)|>
  stack()

# EDA histogram of distribution of Buchanan Votes and Bush Votes
ggplot(tidy_election_wo_pb, aes(x=values, fill = ind)) +
  geom_histogram(bins = 20)+
  theme(legend.position="top")+
  labs(x = "Votes", y= "Count", fill = "Candidates")+
  scale_fill_manual(values=c("#A020F0", "#FF0000")) + 
  facet_wrap(~ind)

# EDA histogram of distribution of Buchanan Votes and Bush Votes under a natural logarithmic transformation
ggplot(tidy_election_wo_pb, aes(x=log(values), fill = ind)) +
  geom_histogram(bins = 20)+
  theme(legend.position="top")+
  labs(x = "log(Votes)", y= "Count", fill = "Candidates")+
  scale_fill_manual(values=c("#A020F0", "#FF0000")) + 
  facet_wrap(~ind)

# EDA scatterplot of Buchanan Votes against Bush Votes
election_wo_pb |> ggplot(aes(x = Bush2000, y = Buchanan2000)) + 
  geom_point() + 
  labs(title ="Association Between Votes for Bush and Votes for Buchanan.", x = "Bush Votes", y = "Buchanan Votes")

# EDA scatterplot of Buchanan Votes against Bush Votes under a natural logarithmic transformation
ggplot(election_wo_pb, aes(y = Buchanan2000, x = Bush2000)) + 
  geom_point() + 
  labs(title ="Association Between Natural Logarithmic Transformation Votes for Bush and Votes for Buchanan.", 
       x = "log(Bush Votes)", 
       y = "log(Buchanan Votes)")+
  scale_x_continuous(trans = "log") + 
  scale_y_continuous(trans = "log")

# Fit a linear regression model to predict Buchanan2000 votes based on Bush2000 votes
election_wo_pb_lm <- lm(log(Buchanan2000) ~ log(Bush2000), data = election_wo_pb)

# Representing the regression table as a dataframe (i.e., tidying the summary() output)
election_wo_pb_lm_table <- summary(election_wo_pb_lm)$coefficients

# Creating a nicely formatted table from the dataframe using the kable package
#   You can find more information about this package here: https://haozhu233.github.io/kableExtra/awesome_table_in_pdf.pdf
election_wo_pb_lm_table |> kbl(col.names = c("Estimate", "Standard Error", "t value", "Pr(>|t|)"), align = "c", booktabs = T, linesep="", digits = c(2, 2, 2, 4)) |> kable_classic(full_width = F, latex_options = c("HOLD_position"))

# Extracting Bush total county votes in Palm Beach
election_pb_bush <- election |> 
              filter(County == "Palm Beach")|>
              pull(Bush2000)
election_pb_bush

# Generate predictions using the double-log regression model for new data (palm_beach_data),
new_election <- data.frame(Bush2000 = election_pb_bush)
palm_beach_predictions <- election_wo_pb_lm |> augment(newdata = new_election, 
                          interval = "prediction", 
                          conf.level = 0.95)
palm_beach_predictions
# Providing both the predicted values and a 95% prediction interval

# Exponentiate the predictions (since we applied log transformation during modeling) to return the predicted Buchanan votes in the original scale (non-log scale).
pred_buchanan_double_log <- palm_beach_predictions |>
                           select(.fitted, .lower, .upper) |>
                            exp()

# Display the transformed predictions (original scale of Buchanan votes)
pred_buchanan_double_log 

# Regression Conditions Analysis on Model
election_wo_pb_lm <- lm(Buchanan2000 ~ Bush2000, data = election_wo_pb)
plot(election_wo_pb_lm)

# Regression Conditions Analysis on Model under natural logarithmic transformation
election_wo_pb_lm_log <- lm(log(Buchanan2000) ~ log(Bush2000), data = election_wo_pb)
plot(election_wo_pb_lm_log)
```
